---
title: "Predicting Covid Vaccination Rates"
author: "Edbert Jao"
date: "12/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load required libraries

```{r}
library(fredr);
library(vars);
library(urca);
library(tidyverse);
library(aTSA);
library(tsDyn);
library(lmtest);
library(xts);
library(lubridate);
```

## Load data

```{r}
setwd("C:\\Users\\edber\\OneDrive\\Desktop\\Road to PHD\\Fall 2021 @ Tufts\\Applications of Econometrics\\Final Project\\Predicting");
covid_df = read.csv("owid-covid-data.csv");
```

## Initial Data Cleaning
```{r}
covid_df %>% str();
covid_df = covid_df %>% dplyr::filter(iso_code == "USA");

df = select(covid_df, date, people_fully_vaccinated_per_hundred, new_vaccinations_smoothed, new_deaths_per_million);
df %>% str();

date = df$date[328:(length(df$people_fully_vaccinated_per_hundred))] %>% ts();
vacc_rate = df$people_fully_vaccinated_per_hundred[328:(length(df$people_fully_vaccinated_per_hundred))] %>% ts();
ln_newvacc = log(df$new_vaccinations_smoothed)[328:(length(df$people_fully_vaccinated_per_hundred))] %>% ts();
new_deaths_per_million = df$new_deaths_per_million[328:(length(df$people_fully_vaccinated_per_hundred))] %>% ts();
dft_reference = cbind(vacc_rate, new_deaths_per_million, ln_newvacc) %>% as.data.frame(); # creating a dataframe to reference for in-sample predictions
```


### Stage 1: Predicting in-sample for the past 30 days, using all days for which data is available from 31 days ago and prior.
Start by specifying the model. The 30 most recent days' data are removed. They will be predicted for. Comparing the predicted values and the in-sample values will show the model's effectiveness.
```{r}
vacc_rate = df$people_fully_vaccinated_per_hundred[328:(length(df$people_fully_vaccinated_per_hundred)-30)] %>% ts();
ln_newvacc = log(df$new_vaccinations_smoothed)[328:(length(df$people_fully_vaccinated_per_hundred)-30)] %>% ts();
new_deaths_per_million = df$new_deaths_per_million[328:(length(df$people_fully_vaccinated_per_hundred)-30)] %>% ts();
dft = cbind(vacc_rate, new_deaths_per_million, ln_newvacc) %>% as.data.frame();

grangertest(ln_newvacc~vacc_rate);             # ln_newvacc granger-causes vacc_rate
grangertest(new_deaths_per_million~vacc_rate); # new_deaths_per_million granger-causes vacc_rate

tseries::adf.test(vacc_rate);                  #stationary
tseries::adf.test(ln_newvacc);                 #non-stationary
tseries::adf.test(new_deaths_per_million);     #non-stationary

cbind(vacc_rate, new_deaths_per_million, ln_newvacc) %>% autoplot() + xlab("Days since 2020-12-14");

lagselect = VARselect(dft);
lagselect;                                     # AIC recommends 10 lags

ctest1e = ca.jo(dft, type = "eigen", K = 10);  # Johansen procedure: testing for and estimating cointegrating systems
ctest1e %>% summary();                         # strong evidence for 2 cointegrating relationship: VECM is viable
                                               # all eigenvalues lie within the unit circle
```

## Diagnostic tests and plots for the model.
```{r}
vecm1 = VECM(dft, 10, r = 2, estim = "ML");
vecm1 %>% summary();

model1var <- vec2var(ctest1e, r = 2);
serial.test(model1var, type = "BG");           # strong evidence of serial correlation 
vars::arch.test(model1var);                    # strong evidence of heteroskedastic residuals
normality.test(model1var);                     # strong evidence residuals are not normally distributed

plot(irf(model1var, impulse = "new_deaths_per_million", response = "vacc_rate", n.ahead = 30, boot = TRUE));
plot(irf(model1var, impulse = "ln_newvacc", response = "vacc_rate", n.ahead = 30, boot = TRUE));

plot(fevd(model1var));
```

## Forecast the last 30 days vaccination rate using the model. Then calculate in sample Mean Squared Prediction Error
```{r}
forecast1 <- predict(model1var, n.ahead = 30, ci = 0.95);
fanchart(forecast1, names = "vacc_rate", main = "Fanchart for Vaccination Rate", xlab = "Days since 2020-12-14", ylab = "vacc_rate");  
forecast1$fcst$vacc_rate[,1];

in_sampleMSPE = (1/30) * sum((forecast1$fcst$vacc_rate[,1]%>%as.matrix()) - (dft_reference$vacc_rate[343:372]%>%as.matrix()))^2;
in_sampleMSPE;
```

### Stage 2: Predicting out-of-sample, for the next 30 days. Diagnostic tests return the same conclusions even after returning 30 days to the sample.

```{r}
vacc_rate2 = df$people_fully_vaccinated_per_hundred[328:(length(df$people_fully_vaccinated_per_hundred))] %>% ts();
ln_newvacc2 = log(df$new_vaccinations_smoothed)[328:(length(df$people_fully_vaccinated_per_hundred))] %>% ts();
new_deaths_per_million2 = df$new_deaths_per_million[328:(length(df$people_fully_vaccinated_per_hundred))] %>% ts();
dft2 = cbind(vacc_rate2, new_deaths_per_million2, ln_newvacc2) %>% as.data.frame();

grangertest(ln_newvacc2~vacc_rate2);             # ln_newvacc granger-causes vacc_rate
grangertest(new_deaths_per_million2~vacc_rate2); # new_deaths_per_million granger-causes vacc_rate

tseries::adf.test(vacc_rate2);                  #stationary
tseries::adf.test(ln_newvacc2);                 #non-stationary
tseries::adf.test(new_deaths_per_million2);     #non-stationary
lagselect2 = VARselect(dft2);
lagselect2;                                     # AIC recommends 10 lags

ctest2e = ca.jo(dft2, type = "eigen", K = 10);  # Johansen procedure: testing for and estimating cointegrating systems
ctest2e %>% summary();                          # strong evidence for 2 cointegrating relationship: VECM is viable
                                                # all eigenvalues lie within the unit circle

vecm2 = VECM(dft2, 10, r = 2, estim = "ML");
vecm2 %>% summary();

model2var <- vec2var(ctest2e, r = 2);
serial.test(model2var, type = "BG");           # strong evidence of serial correlation 
vars::arch.test(model2var);                    # strong evidence of heteroskedastic residuals
normality.test(model2var);                     # strong evidence residuals are not normally distributed

```
### Stage 2 Diagnostic Plots highly resemble Stage 1 Diagnostic Plots
```{r}
plot(irf(model2var, impulse = "new_deaths_per_million2", response = "vacc_rate2", n.ahead = 30, boot = TRUE));
plot(irf(model2var, impulse = "ln_newvacc2", response = "vacc_rate2", n.ahead = 30, boot = TRUE));

plot(fevd(model2var));
```

## Forecast vaccination rate for the next 30 days.
```{r}
forecast2 <- predict(model2var, n.ahead = 30, ci = 0.95);
fanchart(forecast2, names = "vacc_rate2", main = "Fanchart for Vaccination Rate", xlab = "Days since 2020-12-14", ylab = "vacc_rate");

forecast2$fcst$vacc_rate2[,1];
```

